---
---
title: "Interaction Regression: BMI and Age Effects on Blood Pressure"
author: "Brett Anwar"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
library(tidyverse)

df <- read_csv("data/processed/nhanes_clean.csv")
head(df)
```
```{r}
model_interaction <- lm(bp ~ bmi * age + smoking, data = df)
summary(model_interaction)
```
```{r}
bmi_seq <- seq(min(df$bmi), max(df$bmi), length.out = 200)

pred_df <- expand.grid(
  bmi = bmi_seq,
  age = c(25, 45, 65),
  smoking = 0
)

pred_df$bp_hat <- predict(model_interaction, newdata = pred_df)

ggplot(df, aes(x = bmi, y = bp)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_line(data = pred_df, aes(x = bmi, y = bp_hat, color = factor(age)), linewidth = 1) +
  labs(
    title = "Interaction Regression: BMI vs Systolic BP",
    subtitle = "Lines show fitted values at ages 25, 45, and 65 (smoking held constant)",
    x = "BMI",
    y = "Systolic Blood Pressure (mmHg)",
    color = "Age"
  )
```

- **BMI × Age**: Based on the model result, the interaction between BMI and age has a significant relation with blood pressure (p < 0.001).
- **BMI**: Based on the model result, BMI has a significant relation with blood pressure (estimate = 0.836, p < 0.001).
- **Age**: Based on the model result, age has a significant relation with blood pressure (estimate = 0.834, p < 0.001).
- **Smoking**: Based on the model result, smoking has no significant relation with blood pressure (p = 0.542).
- **Direction of Interaction**: Based on the model result, the interaction coefficient is negative (estimate = -0.0112), which suggests that the effect of BMI on blood pressure becomes smaller as age increases.
- **Adjusted R²**: 0.2622

Overall, this model suggests that BMI is positively associated with systolic blood pressure, but the strength of this relationship depends on age. Compared to the previous model, this specification provides a modest improvement in explanatory power and captures a more flexible relationship between BMI, age, and blood pressure.

