---
title: "Does Body Mass Index Predict Blood Pressure? Evidence from NHANES"
author: "Your Name"
date: "`r Sys.Date()`"
output: pdf_document
---

This notebook loads the NHANES data for our STAT203 project and prepares the datasets for later cleaning, merging, and regression analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load Required Packages

```{r}
library(tidyverse)
library(haven)
library(ggplot2)
library(stargazer)
```

## 2. Load NHANES Data

NHANES stores different types of information in separate files. For this project, we will use the demographics file, the body measures file, and the blood pressure file. Each dataset contains information on the same individuals, and later we will merge them using the participant identifier `SEQN`.

```{r}
data_path <- "../data/external/"

demo <- read_xpt(file.path(data_path, "DEMO_J.xpt"))
bmx  <- read_xpt(file.path(data_path, "BMX_J.xpt"))
bpx  <- read_xpt(file.path(data_path, "BPX_J.xpt"))
smq  <- read_xpt(file.path(data_path, "SMQ_J.xpt"))
```

## 3. Inspect the Data

At this stage, the goal is simply to confirm that the files loaded correctly and to get familiar with the structure of each dataset.

```{r}
head(demo)
head(bmx)
head(bpx)
head(smq)
```

```{r}
cat("Demographic Data Columns...\n\n")
colnames(demo)
cat("\n")
cat("Body Measure Data Columns...\n\n")
colnames(bmx)
cat("\n")
cat("Blood Pressure Data Columns...\n\n")
colnames(bpx)
cat("\n")
cat("Smoking Questioner Data Columns...\n\n")
colnames(smq)
```

```{r}
demo_small <- demo %>%
  select(SEQN, RIDAGEYR)

bmx_small <- bmx %>%
  select(SEQN, BMXBMI)

bpx_small <- bpx %>%
  select(SEQN, BPXSY1)

smq_small <- smq %>%
  select(SEQN, SMQ020)
```

```{r}
df <- demo_small %>%
  inner_join(bmx_small, by = "SEQN") %>%
  inner_join(bpx_small, by = "SEQN") %>%
  inner_join(smq_small, by = "SEQN")
```

```{r}
df
unique(df$SMQ020)
```

```{r}
df_clean <- df %>%
  drop_na() %>%
  mutate(
    smoking = ifelse(SMQ020 == 1, 1,
                     ifelse(SMQ020 == 2, 0, NA))
  ) %>%
  drop_na(smoking) %>%
  select(SEQN, RIDAGEYR, BMXBMI, BPXSY1, smoking)
df_clean
```

```{r}
df_clean <- df_clean %>%
  rename(
    age = RIDAGEYR,
    bmi = BMXBMI,
    bp = BPXSY1
  )

df_clean
```

```{r}
write_csv(df_clean, "../data/processed/nhanes_clean.csv")
```

```{r}
ggplot(df_clean, aes(x = bmi, y = bp)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "BMI vs Blood Pressure",
    x = "Body Mass Index (BMI)",
    y = "Systolic Blood Pressure"
  )
```

As BMI Increases --> Blood Pressure tends to increase

While BMI shows a positive association with systolic blood pressure, there is substantial variability, suggesting that additional factors (e.g., age, smoking) play an important role in explaining blood pressure differences.

```{r}
ggplot(df_clean, aes(x = bmi, y = bp)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "BMI vs Blood Pressure with Regression Line",
    x = "BMI",
    y = "Systolic Blood Pressure"
  )
```

## Model1: Simple Linear Regression Model
On average, how does blood pressure change as BMI changes?
$$
bp = \beta_0 + \beta_1 bmi + \epsilon
$$
### Model Summary:

1. Coefficient for BMI
A one-unit increase in BMI is associated with an average increase of about 0.36 units in systolic blood pressure.

2. Statistical significance: P-Value is very small
There is very strong statistical evidence that BMI is positively associated with blood pressure.

3. R-squared
BMI explains only about 1.8% of the variation in blood pressure.

4. Residual standard error
On average, predictions are off by about ±19 units of blood pressure

While BMI shows a statistically significant positive association with blood pressure, it explains only a very small portion of the variation, suggesting that other factors play a substantial role.
```{r}
model1 <- lm(bp ~ bmi, data = df_clean)

summary(model1)
```
### Model2: Multiple Linear Regression
$$
bp = \beta_0 + \beta_1 bmi + \beta_2 age + \beta_3 smoking + \epsilon
$$

### Model2 Summary:
What is the relationship between BMI and blood pressure holding age and smoking constant?

1. BMI coefficient
Holding age and smoking constant, a one-unit increase in BMI is associated with a 0.31 increase in blood pressure.
Note: Part of the relationship in Model 1 was actually due to confounding (mainly age).

2. Age coefficient
Each additional year of age is associated with about a 0.51 increase in blood pressure, holding BMI and smoking constant.

3. Smoking coefficient
Smoking does not show a statistically significant association with blood pressure in this model.

4. R-squared
The model now explains a much larger portion of the variation in blood pressure.

5. Residual standard error
Improved from 19.3 to 16.8, hence, Predictions are more accurate now.

Conclusion:

The simple regression model shows a statistically significant positive relationship between BMI and systolic blood pressure. However, the model explains only a small fraction of the variation in blood pressure.

After including additional variables such as age and smoking status, the explanatory power of the model increases substantially. Age emerges as a strong and significant predictor, while the effect of BMI remains positive but slightly reduced, suggesting that part of the initial association was influenced by confounding factors.

Smoking does not appear to have a statistically significant relationship with blood pressure in this specification.

```{r}
model2 <- lm(bp ~ bmi + age + smoking, data = df_clean)

summary(model2)
```


### Regression Table

```{r}
stargazer(model1, model2,
          type = "text",
          title = "Regression Results: BMI and Blood Pressure",
          dep.var.labels = "Systolic Blood Pressure",
          covariate.labels = c("BMI", "Age", "Smoking"),
          digits = 3)
```

```{r}
residuals_model2 <- resid(model2)
fitted_model2 <- fitted(model2)
```

```{r}
ggplot(data.frame(fitted_model2, residuals_model2), 
       aes(x = fitted_model2, y = residuals_model2)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs Fitted",
       x = "Fitted Values",
       y = "Residuals")
```

### Histogram of Residulas

```{r}
ggplot(data.frame(residuals_model2), aes(x = residuals_model2)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  labs(title = "Histogram of Residuals")
```

### Q-Q Plot

```{r}
qqnorm(residuals_model2)
qqline(residuals_model2, col = "red")
```

### Regression Assumptions

The assumptions of the linear regression model were evaluated using residual diagnostic plots. The residuals versus fitted values plot shows that residuals are generally centered around zero without a clear systematic pattern, suggesting that the linearity assumption is reasonably satisfied. However, there is some evidence of increasing variability in residuals at higher fitted values, indicating mild heteroskedasticity.

The histogram of residuals shows an approximately bell-shaped distribution centered around zero, although with slight right skewness. This is further supported by the Q-Q plot, where most observations lie close to the reference line, with some deviation in the upper tail. These results suggest that the normality assumption is largely satisfied, with minor deviations at the extremes.

Overall, while there are small departures from ideal assumptions, particularly in terms of constant variance, these deviations are not severe and do not undermine the validity of the linear regression model for descriptive analysis.