---
title: "Exploring New Variables for BMI-Blood Pressure Model"
author: "Ronald Tieu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Purpose

This notebook explores whether adding new NHANES variables improves the
BMI-blood pressure regression model. We test: sex, race/ethnicity, BP
medication, physical activity, alcohol use, sodium intake, and an upgraded
3-level smoking variable. We also switch from a single BP reading to the
average of available readings.

## 1. Load Data

We start from the cleaned v2 dataset (averaged BP, sex, race, 3-level smoking),
then join in the additional NHANES files for physical activity, alcohol, sodium,
and BP medication to test whether they improve the model.

```{r}
library(tidyverse)
library(haven)
library(stargazer)

# Base dataset from cleaned v2
df_base <- read_csv("../data/processed/nhanes_clean_v2.csv")

# Additional NHANES files for extra variables
data_path <- "../data/external/"
paq <- read_xpt(file.path(data_path, "PAQ_J.xpt"))
alq <- read_xpt(file.path(data_path, "ALQ_J.xpt"))
dr1 <- read_xpt(file.path(data_path, "DR1TOT_J.xpt"))
bpq <- read_xpt(file.path(data_path, "BPQ_J.xpt"))
```

## 2. Merge Extra Variables

```{r}
df <- df_base %>%
  inner_join(paq %>% select(SEQN, PAQ605, PAQ650), by = "SEQN") %>%
  inner_join(alq %>% select(SEQN, ALQ130), by = "SEQN") %>%
  inner_join(dr1 %>% select(SEQN, DR1TSODI), by = "SEQN") %>%
  inner_join(bpq %>% select(SEQN, BPQ050A), by = "SEQN")

cat("Merged dataset:", nrow(df), "rows\n")
```

## 3. Clean and Recode Extra Variables

```{r}
df_clean <- df %>%
  mutate(
    # Rename sex for consistency with this notebook
    female = sex,

    # Race/ethnicity as factor
    race = factor(race,
                  levels = c(3, 1, 2, 4, 6, 7),
                  labels = c("White", "Mex American", "Other Hispanic",
                             "Black", "Asian", "Other/Multi")),

    # Smoking already 3-level in v2, just factor it
    smoking = factor(smoking, levels = c("never", "former", "current")),

    # BP medication: 1 = yes, 0 = no
    bp_meds = case_when(
      BPQ050A == 1 ~ 1,
      BPQ050A == 2 ~ 0,
      TRUE ~ NA_real_
    ),

    # Physical activity: 1 = any recreational activity
    active = case_when(
      PAQ605 %in% c(7, 9) ~ NA_real_,
      PAQ650 %in% c(7, 9) ~ NA_real_,
      PAQ605 == 1 | PAQ650 == 1 ~ 1,
      TRUE ~ 0
    ),

    # Alcohol: avg drinks/day (777/999 = refused/don't know)
    alcohol = ifelse(ALQ130 %in% c(777, 999), NA, ALQ130),

    # Sodium: total mg
    sodium = DR1TSODI
  ) %>%
  select(SEQN, bp, bmi, age, female, race, smoking, bp_meds, active,
         alcohol, sodium)

cat("After recoding:", nrow(df_clean), "rows\n")
```

## 4. Missingness Summary

```{r}
cat("=== Missingness per variable ===\n\n")
df_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(pct_missing = round(100 * n_missing / nrow(df_clean), 1)) %>%
  print(n = 20)
```

```{r}
# Complete cases for the full model
df_full <- df_clean %>% drop_na()
cat("\nComplete cases (all variables):", nrow(df_full), "out of", nrow(df_clean), "\n")
cat("Dropped:", nrow(df_clean) - nrow(df_full), "observations\n")
```

## 5. Summary Statistics

```{r}
cat("=== Continuous variables ===\n")
df_full %>%
  select(bp, bmi, age, alcohol, sodium) %>%
  summary()

cat("\n=== Binary/Categorical variables ===\n")
cat("\nSex:\n"); table(df_full$female)
cat("\nRace:\n"); table(df_full$race)
cat("\nSmoking:\n"); table(df_full$smoking)
cat("\nBP Medication:\n"); table(df_full$bp_meds)
cat("\nPhysically Active:\n"); table(df_full$active)
```

## 6. Model Comparison

We build models incrementally to see which variables improve fit.

```{r}
m1 <- lm(bp ~ bmi, data = df_full)
m2 <- lm(bp ~ bmi + age + smoking, data = df_full)
m3 <- lm(bp ~ bmi + age + smoking + female + race + bp_meds, data = df_full)
m4 <- lm(bp ~ bmi + age + smoking + female + race + bp_meds + active, data = df_full)
m5 <- lm(bp ~ bmi + age + smoking + female + race + bp_meds + active + alcohol, data = df_full)
m6 <- lm(bp ~ bmi + age + smoking + female + race + bp_meds + active + alcohol + sodium, data = df_full)
m7 <- lm(bp ~ bmi * age + smoking + female + race + bp_meds + I(age^2), data = df_full)
```

```{r results='asis'}
stargazer(
  m1, m2, m3, m4, m5, m6, m7,
  type = "text",
  title = "Variable Exploration: Incremental Model Comparison",
  dep.var.labels = "Systolic Blood Pressure (mmHg)",
  digits = 3,
  omit.stat = c("f"),
  column.labels = c("Baseline", "+Age/Smk", "+Sex/Race/Meds",
                     "+Active", "+Alcohol", "+Sodium", "Interaction"),
  model.numbers = TRUE
)
```

## 7. Fit Comparison Table

```{r}
models <- list(m1, m2, m3, m4, m5, m6, m7)
model_names <- c("M1: BMI only",
                 "M2: + age + smoking(3)",
                 "M3: + female + race + bp_meds",
                 "M4: + active",
                 "M5: + alcohol",
                 "M6: + sodium",
                 "M7: interaction + age^2")

comparison <- tibble(
  Model = model_names,
  R2 = sapply(models, function(m) round(summary(m)$r.squared, 4)),
  Adj_R2 = sapply(models, function(m) round(summary(m)$adj.r.squared, 4)),
  AIC = sapply(models, function(m) round(AIC(m), 1)),
  BIC = sapply(models, function(m) round(BIC(m), 1)),
  RSE = sapply(models, function(m) round(sigma(m), 2)),
  n = sapply(models, function(m) nobs(m))
)

print(comparison, n = 10, width = Inf)
```

## 8. Q-Q Plot of Model 5 Residuals

The final report notes that residuals are right-skewed. This Q-Q plot confirms it: the upper quantiles pull away from the reference line, indicating heavier right tails than a normal distribution.

```{r fig.width=5, fig.height=4}
library(here)

# Rebuild the confirmation set to match the report
data_path <- here("data/external/")
demo <- read_xpt(file.path(data_path, "DEMO_J.xpt"))
bmx  <- read_xpt(file.path(data_path, "BMX_J.xpt"))
bpx  <- read_xpt(file.path(data_path, "BPX_J.xpt"))
smq  <- read_xpt(file.path(data_path, "SMQ_J.xpt"))

df_raw <- demo %>% select(SEQN, RIDAGEYR, RIAGENDR, RIDRETH3) %>%
  inner_join(bmx %>% select(SEQN, BMXBMI), by = "SEQN") %>%
  inner_join(bpx %>% select(SEQN, BPXSY1, BPXSY2, BPXSY3), by = "SEQN") %>%
  inner_join(smq %>% select(SEQN, SMQ020, SMQ040), by = "SEQN")

df_raw <- df_raw %>%
  mutate(
    bp = rowMeans(select(., BPXSY1, BPXSY2, BPXSY3), na.rm = TRUE),
    bp = ifelse(is.nan(bp), NA, bp),
    bmi = BMXBMI, age = RIDAGEYR,
    sex = ifelse(RIAGENDR == 2, 1, 0),
    race = factor(RIDRETH3, levels = c(3,1,2,4,6,7),
                  labels = c("White","Mex American","Other Hispanic","Black","Asian","Other/Multi")),
    smoking = case_when(
      SMQ020 == 2 ~ "never", SMQ020 == 1 & SMQ040 == 3 ~ "former",
      SMQ020 == 1 & SMQ040 %in% c(1,2) ~ "current", TRUE ~ NA_character_),
    smoking = factor(smoking, levels = c("never","former","current"))
  ) %>% select(SEQN, bp, bmi, age, sex, race, smoking) %>% drop_na()

set.seed(42)
split_idx <- sample(seq_len(nrow(df_raw)), size = floor(nrow(df_raw) * 0.30))
df_confirm <- df_raw[-split_idx, ]

model_five <- lm(bp ~ bmi * age + smoking + sex + race + I(age^2), data = df_confirm)

qqnorm(resid(model_five), main = "Q-Q Plot: Model 5 Residuals (Confirmation Set)",
       pch = 20, cex = 0.5, col = "#2C7FB8")
qqline(resid(model_five), col = "#D95F0E", lwd = 2)
```

## 9. Key Takeaways

Based on the model comparison above, we can determine:

- Which variables meaningfully increase R-squared
- Which variables are statistically significant
- Whether the sample size drops too much when including all variables
- Whether alcohol and sodium are worth the added complexity
- How the 3-level smoking variable compares to the old binary version
