---
title: "Logistic Regression: Predicting Hypertension from BMI, Age, and Smoking"
author: "Ronald Tieu"
date: "`r Sys.Date()`"
output: pdf_document
---

## Overview

In this analysis, we will use logistic regression to predict whether a person is hypertensive based on their BMI, age, and smoking status.

A person is classified as hypertensive if their systolic blood pressure is 130 mmHg or higher, based on the American Heart Association guideline.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages and Data

```{r}
library(tidyverse)

df <- read_csv("../data/processed/nhanes_clean.csv")
head(df)
```

## Create the Hypertension Variable

We create a binary variable: 1 if systolic blood pressure is \>= 130, 0 otherwise.

```{r}
df <- df %>%
  mutate(hypertension = ifelse(bp >= 130, 1, 0))

table(df$hypertension)
prop.table(table(df$hypertension))
```

## Explore the Data

Before modeling, we look at how hypertension rates differ across the predictor variables. The results show that BMI does have some relations but not a clean separation.

```{r}
ggplot(df, aes(x = bmi, fill = factor(hypertension, labels = c("No", "Yes")))) +
  geom_density(alpha = 0.5) +
  labs(title = "BMI Distribution by Hypertension Status",
       x = "BMI", fill = "High Blood Pressure")
```

This bar graph shows that hypertension rates does increase, with how the youngest group has the lowest rates, while the oldest is more than 3x higher.

```{r}
df %>%
  mutate(age_group = cut(age, breaks = c(0, 30, 45, 60, Inf),
                         labels = c("18-30", "31-45", "46-60", "60+"))) %>%
  group_by(age_group) %>%
  summarise(hypertension_rate = mean(hypertension)) %>%
  ggplot(aes(x = age_group, y = hypertension_rate, fill = age_group)) +
  geom_col() +
  labs(title = "Hypertension Rate by Age Group",
       x = "Age Group", y = "Hypertension Rate") +
  scale_fill_brewer(guide = "none")
```

We see two things here:

-   Out of 2,857 non smokers, the hypertension rate is 33.3%

-   Out of 1,9676 smokers, the hypertension rate is 40.3%

```{r}
df %>%
  group_by(smoking) %>%
  summarise(
    count = n(),
    hypertension_rate = mean(hypertension)
  )
```

## Fit the Logistic Regression Model

We fit a logistic regression model using BMI, age, and smoking as predictors.

```{r}
model <- glm(hypertension ~ bmi + age + smoking, data = df, family = binomial)
summary(model)
```

## Interpret the Coefficients (Odds Ratios)

Odds ratios tell us how much the odds of hypertension change for a one-unit increase in each predictor.

-   Odds ratio \> 1: increases the odds of hypertension
-   Odds ratio \< 1: decreases the odds of hypertension

We can see that bmi, age, and smoking is greater than 1. So they are all positive predictors of hypertension.

```{r}
odds_ratios <- exp(coef(model))
ci <- exp(confint.default(model))

data.frame(
    Odds_Ratio = round(odds_ratios, 3),
    CI_Lower = round(ci[, 1], 3),
    CI_Upper = round(ci[, 2], 3)
)
```

## Make Predictions and Evaluate the Model

Using 0.5 as the cutoff point. If the model predicted is \>= 0.5, then we will classify the person as hypertensive.

```{r}
df$predicted_prob <- predict(model, type = "response")
df$predicted_class <- ifelse(df$predicted_prob >= 0.5, 1, 0)
```

### Confusion Matrix

-   2526: predicted not hypertensive, actually not hypertensive (correct negatives)
-   928: predicted hypertensive, actually hypertensive (correct positives)
-   560: predicted hypertensive, actually not hypertensive (false positives)
-   819: predicted not hypertensive, actually hypertensive (false negatives)

```{r}
confusion <- table(Predicted = df$predicted_class, Actual = df$hypertension)
confusion
```

### Accuracy

The model correctly classified 71.5% of individuals as either hypertensive or non-hypertensive correctly.

```{r}
accuracy <- sum(diag(confusion)) / sum(confusion)
cat("Accuracy:", round(accuracy, 3), "\n")
```

## ROC Curve

The ROC curve shows model performance across all possible thresholds, which is why the values are different between the accuracy and the AUC (area under the curve), which summarizes overall performance.

```{r}
library(pROC)

roc_obj <- roc(df$hypertension, df$predicted_prob)
auc_val <- auc(roc_obj)

ggroc(roc_obj, color = "tomato", linewidth = 1) +
  labs(title = paste("ROC Curve (AUC =", round(auc_val, 3), ")"),
       x = "Specificity",
       y = "Sensitivity") +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray")
```

The logistical regression results show that age and BMI were significant positive predictors of hypertension. As age and BMI increased, the odds of being hypertensive also increased about 6.2% for each additional year of age and 3.6% for each one unit increase in BMI.

Something slightly shocking was that smoking was not statistically significant after accounting for age and BMI, with the p-value of 0.963, which is far above 0.05. Overall the model had a moderate predictive performance, when an accuracy rate of 71.5% and an AUC of 0.774.
