---
title: "Polynomial Regression: Nonlinear Effect of BMI on Blood Pressure"
author: "Edward"
date: "`r Sys.Date()`"
output: pdf_document
---

## Overview

In this analysis, I applied polynomial regression to look at how BMI relates to systolic blood pressure. I wanted to see if the relationship between BMI and BP is really a straight line, or if it curves — for example, whether a higher BMI has a bigger impact on BP than a lower one. To test this, I added a squared BMI term (BMI²) to the model. I also included age and smoking as additional predictors.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load Packages and Data

```{r}
library(tidyverse)

df <- read_csv("../data/processed/nhanes_clean.csv")
head(df)
```

## 2. Fit the Polynomial Model

`poly(bmi, 2)` adds both a linear and quadratic BMI term, using orthogonal polynomials to reduce multicollinearity.

```{r}
model_poly <- lm(bp ~ poly(bmi, 2) + age + smoking, data = df)
summary(model_poly)
```

## 3. Visualize the BMI Curve

I plotted the fitted curve over the raw data to see if the model picks up any curvature. I kept age at its mean and set smoking to 0 just to make the plot easier to read.

```{r}
bmi_seq <- seq(min(df$bmi), max(df$bmi), length.out = 200)

pred_df <- data.frame(
  bmi     = bmi_seq,
  age     = mean(df$age),
  smoking = 0
)
pred_df$bp_hat <- predict(model_poly, newdata = pred_df)

ggplot(df, aes(x = bmi, y = bp)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_line(data = pred_df, aes(x = bmi, y = bp_hat), color = "red", linewidth = 1) +
  labs(
    title = "Polynomial Regression: BMI vs Systolic BP",
    subtitle = "Red curve: quadratic fit (age and smoking held constant)",
    x = "BMI",
    y = "Systolic Blood Pressure (mmHg)"
  )
```

## 4. Interpretation

- **BMI²**: Based on the model result, BMI² has no significant relation with blood pressure (p = 0.268).
- **Age**: Based on the model result, age has a significant relation with blood pressure (p < 0.001).
- **Smoking**: Based on the model result, smoking has no significant relation with blood pressure (p = 0.591).
- **Adjusted R²**: 0.2569
