---
title: "Model 4: Quadratic Regression Extension of the BMI and Blood Pressure Model"
author: "Ronald Tieu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(stargazer)

df <- read_csv("../data/processed/nhanes_clean.csv")

head(df)
summary(df)
```

```{r}
model2 <- lm(bp ~ bmi + age + smoking, data = df)
summary(model2)
```

```{r}
model3 <- lm(bp ~ bmi * age + smoking, data = df)
summary(model3)
```

```{r}
model4 <- lm(bp ~ bmi * age + smoking + I(age^2), data = df)
summary(model4)
```

```{r}
model_comp <- tibble(
  Model = c("Model 2: BMI + Age + Smoking",
            "Model 3: BMI x Age Interaction + Smoking",
            "Model 4: BMI x Age + Age^2 + Smoking"),
  R2 = c(
    summary(model2)$r.squared,
    summary(model3)$r.squared,
    summary(model4)$r.squared
  ),
  Adj_R2 = c(
    summary(model2)$adj.r.squared,
    summary(model3)$adj.r.squared,
    summary(model4)$adj.r.squared
  ),
  AIC = c(
    AIC(model2),
    AIC(model3),
    AIC(model4)
  ),
  BIC = c(
    BIC(model2),
    BIC(model3),
    BIC(model4)
  ),
  RSE = c(
    sigma(model2),
    sigma(model3),
    sigma(model4)
  )
)

model_comp
```

```{r}
anova(model2, model3)
```

```{r}
anova(model3, model4)
```

```{r}
stargazer(
  model2, model3, model4,
  type = "text",
  title = "Regression Results: Blood Pressure Models",
  dep.var.labels = "Systolic Blood Pressure",
  digits = 3
)
```

22 - healthy bmi

30 - moderate

30+ - obese

```{r}
bmi_vals <- c(22, 30, 38)

plot_df <- expand.grid(
  age = seq(min(df$age, na.rm = TRUE), max(df$age, na.rm = TRUE), length.out = 200),
  bmi = bmi_vals,
  smoking = 0
)

plot_df$bp_hat <- predict(model4, newdata = plot_df)
plot_df$bmi_group <- factor(plot_df$bmi)

ggplot(df, aes(x = age, y = bp)) +
  geom_point(alpha = 0.12, color = "steelblue") +
  geom_line(
    data = plot_df,
    aes(x = age, y = bp_hat, color = bmi_group),
    linewidth = 1.2
  ) +
  labs(
    title = "Quadratic Age Model: Blood Pressure by Age",
    subtitle = "Predicted blood pressure for BMI 22, 30, and 38 (smoking held constant at 0)",
    x = "Age",
    y = "Systolic Blood Pressure",
    color = "BMI"
  ) +
  theme_minimal()
```

Ali's multiple linear regression model assumes constant effects of BMI and age, while Brett's interaction model allows the BMI-blood pressure relationship to vary with age.

Model 4 extends Brett's specification by testing whether blood pressure rises with age at a constant rate or accelerates at older ages.

The negative `bmi:age` coefficient remains significant, indicating that BMI has a smaller association with blood pressure among older adults. The positive and significant `age^2` term suggests that the age-blood pressure relationship steepens at older ages.

Model fit improves modestly across the sequence, with adjusted `R^2` increasing from `r round(summary(model2)$adj.r.squared, 4)` to `r round(summary(model3)$adj.r.squared, 4)` to `r round(summary(model4)$adj.r.squared, 4)`.
