---
title: "Test: Adding Sex, Race, and 3-Level Smoking to Original Models"
author: "Ronald Tieu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Purpose

Test whether adding sex, race/ethnicity, and upgraded 3-level smoking improves the original models — without losing sample size.

## 1. Build Dataset

```{r}
library(tidyverse)
library(stargazer)

# Load from the cleaned v2 dataset (averaged BP, sex, race, 3-level smoking)
df_v2 <- read_csv("../data/processed/nhanes_clean_v2.csv")

# Recode for modeling
df_clean <- df_v2 %>%
  mutate(
    race = factor(race,
                  levels = c(3, 1, 2, 4, 6, 7),
                  labels = c("White", "Mex American", "Other Hispanic",
                             "Black", "Asian", "Other/Multi")),
    smoking3 = factor(smoking, levels = c("never", "former", "current")),
    # Keep binary smoking for old vs new comparison
    smoking_bin = smoking_bin
  )

cat("Final sample size:", nrow(df_clean), "\n")
cat("Age range:", min(df_clean$age), "-", max(df_clean$age), "\n")
cat("Mean age:", round(mean(df_clean$age), 1), "\n")
```

## 2. Exploratory Data Analysis

### Summary Statistics

```{r}
cat("=== Continuous Variables ===\n")
df_clean %>%
  select(bp, bmi, age) %>%
  summary()

cat("\n=== Categorical Variables ===\n")
cat("\nSex (0=Male, 1=Female):\n")
table(df_clean$sex) %>% print()
cat("\nRace/Ethnicity:\n")
table(df_clean$race) %>% print()
cat("\nSmoking Status:\n")
table(df_clean$smoking3) %>% print()
```

### Outcome Variable: Systolic Blood Pressure Distribution

```{r, fig.width=6, fig.height=3.5}
ggplot(df_clean, aes(x = bp)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7, color = "white") +
  geom_vline(xintercept = mean(df_clean$bp), color = "red", linetype = "dashed", linewidth = 0.8) +
  labs(title = "Distribution of Systolic Blood Pressure",
       subtitle = "Red dashed line = mean",
       x = "Systolic Blood Pressure (mmHg)", y = "Count") +
  theme_minimal()
```

### Predictor Distributions: BMI and Age

```{r, fig.width=8, fig.height=3.5}
p_bmi <- ggplot(df_clean, aes(x = bmi)) +
  geom_histogram(bins = 40, fill = "darkorange", alpha = 0.7, color = "white") +
  labs(title = "Distribution of BMI", x = "BMI", y = "Count") +
  theme_minimal()

p_age <- ggplot(df_clean, aes(x = age)) +
  geom_histogram(bins = 40, fill = "darkgreen", alpha = 0.7, color = "white") +
  labs(title = "Distribution of Age", x = "Age (years)", y = "Count") +
  theme_minimal()

library(patchwork)
p_bmi | p_age
```

### Bivariate: BP vs BMI and BP vs Age

```{r, fig.width=8, fig.height=3.5}
p_bmi_bp <- ggplot(df_clean, aes(x = bmi, y = bp)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "BP vs BMI", x = "BMI", y = "Systolic BP (mmHg)") +
  theme_minimal()

p_age_bp <- ggplot(df_clean, aes(x = age, y = bp)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "BP vs Age", x = "Age (years)", y = "Systolic BP (mmHg)") +
  theme_minimal()

p_bmi_bp | p_age_bp
```

### BP by Sex

```{r, fig.width=5, fig.height=3.5}
ggplot(df_clean, aes(x = factor(sex, labels = c("Male", "Female")), y = bp)) +
  geom_boxplot(fill = c("#4A90D9", "#D94A7A"), alpha = 0.7) +
  labs(title = "Systolic Blood Pressure by Sex",
       x = "Sex", y = "Systolic BP (mmHg)") +
  theme_minimal()
```

### BP by Race/Ethnicity

```{r, fig.width=7, fig.height=4}
ggplot(df_clean, aes(x = reorder(race, bp, FUN = median), y = bp, fill = race)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Systolic Blood Pressure by Race/Ethnicity",
       x = "Race/Ethnicity", y = "Systolic BP (mmHg)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 15, hjust = 1))
```

### BP by Smoking Status

```{r, fig.width=5, fig.height=3.5}
ggplot(df_clean, aes(x = smoking3, y = bp, fill = smoking3)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "Systolic Blood Pressure by Smoking Status",
       x = "Smoking Status", y = "Systolic BP (mmHg)") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Correlation Matrix

```{r, fig.width=5, fig.height=4}
cor_data <- df_clean %>%
  select(bp, bmi, age, sex) %>%
  cor()

cor_long <- as.data.frame(as.table(cor_data)) %>%
  rename(Var1 = Var1, Var2 = Var2, Correlation = Freq)

ggplot(cor_long, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  geom_text(aes(label = round(Correlation, 2)), size = 4) +
  scale_fill_gradient2(low = "#D73027", mid = "white", high = "#4575B4", midpoint = 0) +
  labs(title = "Correlation Matrix") +
  theme_minimal() +
  theme(axis.title = element_blank())
```

------------------------------------------------------------------------

## 3. Compare: Original Models vs New Models

```{r}
# Original Models (binary smoking, single BP reading)
old_m1 <- lm(bp ~ bmi, data = df_clean)
old_m2 <- lm(bp ~ bmi + age + smoking_bin, data = df_clean)
old_m3 <- lm(bp ~ bmi + I(bmi^2) + age + smoking_bin, data = df_clean)
old_m4 <- lm(bp ~ bmi * age + smoking_bin, data = df_clean)
old_m5 <- lm(bp ~ bmi * age + smoking_bin + I(age^2), data = df_clean)

# New Models (3-level smoking + sex + race)
new_m1 <- lm(bp ~ bmi, data = df_clean)
new_m2 <- lm(bp ~ bmi + age + smoking3 + sex + race, data = df_clean)
new_m3 <- lm(bp ~ bmi + I(bmi^2) + age + smoking3 + sex + race, data = df_clean)
new_m4 <- lm(bp ~ bmi * age + smoking3 + sex + race, data = df_clean)
new_m5 <- lm(bp ~ bmi * age + smoking3 + sex + race + I(age^2), data = df_clean)
```

## 3. Side-by-Side R² Comparison

```{r}
old_models <- list(old_m1, old_m2, old_m3, old_m4, old_m5)
new_models <- list(new_m1, new_m2, new_m3, new_m4, new_m5)
model_names <- c("M1: BMI only",
                 "M2: + age + smoking",
                 "M3: + BMI²",
                 "M4: BMI × Age interaction",
                 "M5: + Age²")

comparison <- tibble(
  Model = model_names,
  Old_R2 = sapply(old_models, function(m) round(summary(m)$r.squared, 4)),
  Old_AdjR2 = sapply(old_models, function(m) round(summary(m)$adj.r.squared, 4)),
  New_R2 = sapply(new_models, function(m) round(summary(m)$r.squared, 4)),
  New_AdjR2 = sapply(new_models, function(m) round(summary(m)$adj.r.squared, 4)),
  R2_change = sapply(seq_along(old_models), function(i) {
    round(summary(new_models[[i]])$adj.r.squared - summary(old_models[[i]])$adj.r.squared, 4)
  })
)

print(comparison, width = Inf)
```

## 4. Full Coefficient Table for Best New Model

```{r results='asis'}
stargazer(
  old_m2, new_m2, old_m5, new_m5,
  type = "text",
  title = "Old vs New: Key Model Comparisons",
  dep.var.labels = "Systolic Blood Pressure (mmHg)",
  column.labels = c("Old M2", "New M2", "Old M5", "New M5"),
  digits = 3,
  omit.stat = c("f")
)
```
